---
title: "Getting started with retraction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with retraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`retraction` helps you avoid citing retracted papers. It reads documents and
bibliographies, extracts identifiers, checks them against retraction data, and
reports the retracted references it finds.

The code chunks that reach the network are shown but not run here, so this
article builds offline. Run them in your own session.

```{r setup}
library(retraction)
```

## Checking a file

`check_file()` reads a document or bibliography and checks every reference. It
supports BibTeX and BibLaTeX (`.bib`), CSL-JSON (`.json`), RIS (`.ris`), EndNote
XML and JATS XML (`.xml`), Word (`.docx`), PDF (`.pdf`), and any text document
(`.Rmd`, `.qmd`, `.tex`, `.md`, `.txt`, `.html`), from which DOIs are scraped.

```{r check-file, eval = FALSE}
bib <- system.file("extdata", "example.bib", package = "retraction")
result <- check_file(bib)
result
```

The result is a tibble with one row per reference. Key columns include `status`
(retracted, expression_of_concern, correction, reinstated, none, or unchecked),
`is_retracted` (a convenience flag), `confidence`, `retraction_date`,
`days_since_retraction`, `reason`, and `sources`.

```{r retracted, eval = FALSE}
retracted(result)                 # just the flagged citations
render_report(result, "report.html")
```

## Checking identifiers or a data frame

```{r check-dois, eval = FALSE}
check_dois(c("10.1016/S0140-6736(97)11096-0", "10.1038/s41586-020-2649-2"))
```

The bundled `retraction_example` data frame shows the shape `check_refs()`
expects; columns are auto-detected.

```{r example-data}
retraction_example
```

```{r check-refs, eval = FALSE}
check_refs(retraction_example)
```

## Choosing sources

The default source is Retraction Watch (`"xera"`). You can add Crossref and
OpenAlex, and the results are reconciled across them.

```{r sources, eval = FALSE}
list_backends()
check_dois("10.1016/S0140-6736(97)11096-0",
           sources = c("xera", "crossref", "openalex"))
```

## Offline mode

For bulk or private checking, download a local snapshot once and match against
it. Updates are incremental.

```{r offline, eval = FALSE}
retraction_sync()
check_file("manuscript.Rmd", offline = TRUE)
```

## Identifier normalization

The normalization helpers are exported and useful on their own.

```{r normalize}
normalize_doi("https://doi.org/10.1016/S0140-6736(97)11096-0")
normalize_pmid("PMID: 9500320")
```
